The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics

### 2016-04-12 Papers/InProgress/IJCAI/submission.tex
Add authors
Address reviewer's comments for IJCAI paper

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{Christoph Benzm\"uller\thanks{German Research Foundation DFG \ldots} and Bruno Woltzenlogel Paleo}
\author{}

@@

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in the early 70's, the
  inconsistency of the axioms used in the argument remained unnoticed
  until 2013, when it was detected automatically by the higher-order
  theorem prover \textsc{Leo-II}.  Understanding and verifying the
  refutation generated by the prover turned out to be a time-consuming
  task. Its completion, as reported here, required the
  reconstruction of the refutation in the Isabelle proof assistant, and it
  also led to a novel and more efficient way of automating
  higher-order modal logic \textbf{S5} with a universal accessibility
  relation. Furthermore, the development of an improved syntactical
  hiding for the utilized logic embedding technique allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.
\end{abstract}

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticise, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{sobel2004logic} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf. \cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf. \cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

@@

First successful applications of theorem proving technology in
metaphysics were reported by Oppenheimer and
Zalta~\shortcite{oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on, Rushby~\shortcite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\shortcite{Proslogion} simpler 
and older ontological argument. In contrast, the greater complexity of G\"odel's argument requires the formalization and automation of variants of \emph{higher-order} and \emph{modal} logics.

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz. 
%\cite{ToDo:Leibniz}.  
Although G\"odel's notion of positive property is not exactly the same as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

The modal logic \SFive requires that the accessibility relation be
reflexive, symmetric and transitive. The usual approach to embed
\SFive would be to use the standard translation for \textbf{K} described above
and to state that $r$ is an equivalence relation, e.g., by postulating
the following axioms:
\begin{itemize}
\item Reflexivity: $\forall x. (r~x~x)$
\item Symmetry: $\forall x. \forall y. (r~x~y) \rightarrow (r~y~x)$ 
\item Transitivity: $\forall x. \forall y. \forall z. (r~x~y) \wedge (r~y~z) \rightarrow (r~x~z)$
\end{itemize}
Instead, we consider here a modal logic that we call \SFiveU, which is
characterized by the following condition on $r$:
\begin{itemize}
\item Universality: $\forall x. \forall y. (r~x~y)$
\end{itemize}

It is easy to see that \SFiveU is at least as strong as \SFive:
universality entails that $r$ is an equivalence relation. \SFiveU is,
in fact, \emph{strictly stronger} than \SFive. \SFiveU only admits
complete\footnote{A graph is \emph{complete} iff there is a directed
  edge connecting every ordered pair of vertices.} frames, whereas
\SFive admits non-complete frames as long as all their components are
complete. In other words, in \SFiveU we face one single equivalence
class of possible worlds, while in \SFive we may face several
disconnected equivalence classes. In fact, for this reason, \SFiveU is
considered as metaphysically more appropriate by some philosophers;
cf. \cite[p.~127]{williamson13}.

For \SFiveU, an improved embedding is possible. Universality implies
that the guarding predicates in the definitions of $\Box$ and
$\Diamond$ always hold. Therefore, they can be omitted and the
accessibility relation can be dispensed altogether. The modal
operators can then be defined merely as:
% \begin{itemize}
% \item $\BlackBox \varphi \equiv \forall w. (\varphi w)$ 
% \item $\BlackDia \varphi \equiv \exists w. (\varphi w)$
% \end{itemize}
\[\Box \varphi \equiv \lambda w. \forall v.  \varphi(v) \qquad \text{and}
\qquad \Dia \varphi \equiv \lambda w.\exists v. \varphi(v)\]

The new embedding of \SFiveU in Isabelle/HOL is shown in
Fig.~\ref{QML_S5U}. It is important to note that we 
have  $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$,
despite the differences in the underlying frame structures.

%
% \marginpar{@Bruno: I commented you paragraph here, since I am not
%   convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
%   and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{Necessarily, there exists God}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Wider adoption of HOL theorem proving technology for
reasoning about and within embedded object logics, especially among
non-expert users, is still hindered by the gap between the syntax used
by people, when they write logical formulas with pen and paper, and
the syntax used by HOL theorem provers. Even when the syntax of the
underlying higher-order system is elegant (as is the case in
Isabelle/HOL), the embedding of HOML into HOL may easily expose
details of HOL that may be uncommon to the user, disturbing his/her
experience while using the system. To illustrate this point,
Fig.~\ref{UglyEssence} shows how the definition of essence looked like
in previous work \cite{J28}, where advanced syntax-sugaring features
were not used. It looks notably higher-order, and its style differs
significantly from the common style seen in works on modal logics and
the ontological argument. The following specific issues can be
enumerated:
\begin{enumerate}
\item $\lambda$-abstractions, which are typically a HOL feature, appear explicitly in places where they didn't need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he/she would expect, he/she has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

On the other hand, there is still room for many pragmatical
improvements in Isabelle; just one example: in default setting,
Sledgehammer does not immediately inform the user when a proof has
been found and instead silently first executes a series of
time-consuming proof analysis processes (e.g. its dependency
minimization), before it eventually reports success. For G\"odel's
theorem T3 (\textit{Necessarily, there exists God}), for example, this
phase of silence takes several minutes --- during which the user might
actually give up on the proof attempt --- even though \textsc{Leo-II} already
reported success to Sledgehammer after 2,5 seconds.

@@

%\section*{Acknowledgments} Will be added at a later point.
\paragraph*{Acknowledgments} Will be added at a later point.

>>>>

\title{The Inconsistency in G\"odel's Ontological Argument: \\ A Success Story for AI in Metaphysics}
\author{
  \begin{minipage}{0.5\textwidth}\centering
    Christoph Benzm\"uller\thanks{This work was supported by the 
                                  German National Research Foundation (DFG) under
                                  grants BE 2501/9-1 and BE 2501/11-1.} \\
    Freie Universit\"at Berlin, Germany \\
    c.benzmueller@fu-berlin.com
  \end{minipage}
  \begin{minipage}{0.5\textwidth}\centering
    Bruno Woltzenlogel Paleo \\
    Australian National University, Australia \\
    bruno.wp@gmail.com
  \end{minipage}
}

@@

\begin{abstract}
  This paper discusses the discovery of the inconsistency in G\"odel's
  ontological argument as a success story for artificial
  intelligence. Despite the popularity of the argument since the
  appearance of G\"odel's manuscript in the early 1970's, the
  inconsistency of the axioms used in the argument remained unnoticed
  until 2013, when it was detected automatically by the higher-order
  theorem prover \textsc{Leo-II}.  Understanding and verifying the
  refutation generated by the prover turned out to be a time-consuming
  task. Its completion, as reported here, required the
  reconstruction of the refutation in the Isabelle proof assistant, and it
  also led to a novel and more efficient way of automating
  higher-order modal logic \textbf{S5} with a universal accessibility
  relation. Furthermore, the development of an improved syntactical
  hiding for the utilized logic embedding technique allows the refutation to be
  presented in a human-friendly way, suitable for non-experts in the
  technicalities of higher-order theorem proving. This brings us a
  step closer to wider adoption of logic-based artificial intelligence
  tools by philosophers.
\end{abstract}

\section{Introduction}\label{sec:introduction}
Without exaggeration Kurt G\"{o}del's ontological
argument for the existence of God \cite{GoedelNotes,ScottNotes} is
amongst the most discussed formal proofs in modern literature. A rich
body of publications -- including very recent ones -- present,
discuss, assess, criticize, modify and improve G\"{o}del's original
work (see e.g.~Sobel~\shortcite{sobel2004logic} and Oppy~\shortcite{sep-ontological-arguments} and the
references therein).  In philosophy lectures at universities the
argument is regularly presented as a masterpiece argument in
metaphysics. Since 2013, when Benzm\"uller and Woltzenlogel-Paleo~\shortcite{J30,C40} first
reported their successful initial computer-assisted
analysis of G\"odel's proof and Scott's variant,
their work has received a media repercussion on a global scale\footnote{A
  collection of news articles is available at {\scriptsize
    \url{https://github.com/FormalTheology/GoedelGod/blob/master/Press/LinksToNews.md}}},
and numerous bloggers commented on the proof
\cite{fuhrmann15:_blogg_goedel}.

The in-depth analysis presented here substantially
extends previous computer-assisted studies of G\"odel's ontological
argument. Similarly to the related work \cite{J30,C40} the analysis has
been conducted with automated theorem provers for classical
higher-order logic (HOL; cf.~\cite{andrewsSEP} and the references
therein), even though G\"odel's proof is actually formulated in
higher-order \emph{modal} logic (HOML; cf.~\cite{homl} and the
references therein). To bridge between the two logics we utilise and
further improve the logic embedding approach \cite{J23,C40}, which has
already been employed successfully in preceding related work.

@@

First successful applications of theorem proving technology in
metaphysics were reported by Fitelson, Oppenheimer and
Zalta~\shortcite{FitelsonZalta,oppenheimer11}, who coined the term \textit{Computational Metaphysics} for this new research area and employed the first-order
\textsc{Prover9} \cite{prover9-mace4} in their experiments. Later on, Rushby~\shortcite{rushby13} used the proof assistant \textsc{PVS} \cite{cade92-pvs}. Common to both
works is a significant amount of proof-hand-coding work as well as their
focus on a non-modal formalization of St. Anselm's~\shortcite{Proslogion} simpler 
and older ontological argument. In contrast, the greater complexity of G\"odel's argument requires the formalization and automation of variants of \emph{higher-order} and \emph{modal} logics.

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz~\cite{Adams}.  
Although G\"odel's notion of positive property is not exactly the same as Leibniz's notion of perfection, G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. For instance,
Hazen [\citeyear[p.365]{Hazen}] states that ``G\"odel left this
clause out [\ldots] but this appears to have been an oversight''.
% -- it is
%included in related manuscripts''. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analysis and criticisms of the
argument. Especially since the discovery by Sobel
\shortcite{Sobel} that modal collapse (MC)\footnote{The modal collapse,
  $\phi\rightarrow \Box \phi$, states that contingent truth implies
  necessary truth; it can be interpreted as \textit{everything is
    pre-determined} or even \textit{there is no free will}. } is
entailed by G\"odel's (or also Scott's) axioms, several variants have
been proposed
\cite{Anderson,AndersonGettings,Hajek1,Hajek2,Hajek3,Bjordal}
attempting to avoid the modal collapse. Many of these variants omit
the crucial conjunct in the definition of essence as well.\footnote{As
  these variants also change other axioms, on which the inconsistency
  of G\"odel's axioms depends, it is not necessarily the case that
  these variants are also inconsistent; they must be analyzed
  separately.} Opponents of the argument
(e.g. Oppy [\citeyear[p.226--227]{oppy96:_goedel_ontol_argum};
\citeyear[p.364]{oppy00:_respon_gettin};
\citeyear[p.1068]{oppy08:_higher_order_ontol_argum}]) have also proposed
parodies and other criticisms, referring to variants where the
conjunct is omitted.

@@

The modal logic \SFive requires that the accessibility relation be
reflexive, symmetric and transitive. The usual approach to embed
\SFive would be to use the standard translation for \textbf{K} described above
and to state that $r$ is an equivalence relation, e.g., by postulating
the following axioms:
\begin{itemize}
\item Reflexivity: $\forall x. (r~x~x)$
\item Symmetry: $\forall x. \forall y. (r~x~y) \rightarrow (r~y~x)$ 
\item Transitivity: $\forall x. \forall y. \forall z. (r~x~y) \wedge (r~y~z) \rightarrow (r~x~z)$
\end{itemize}
Instead, we consider here an alternative description, that we call \SFiveU, based on the following condition on $r$:
\begin{itemize}
\item Universality: $\forall x. \forall y. (r~x~y)$
\end{itemize}

It is important to note that $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$ 
\cite{Blackburn}, and therefore \SFive and \SFiveU are traditionally considered to be 
two different descriptions of the same modal logic. Nevertheless, \SFive and \SFiveU 
differ in the shapes of frames they admit:
%It is easy to see that \SFiveU is at least as strong as \SFive:
%universality entails that $r$ is an equivalence relation. \SFiveU is,
%in fact, \emph{strictly stronger} than \SFive. 
\SFiveU only admits
complete\footnote{A graph is \emph{complete} iff there is a directed
  edge connecting every ordered pair of vertices.} frames, whereas
\SFive admits non-complete frames as long as all their components are
complete. In other words, in \SFiveU we face one single equivalence
class of possible worlds, while in \SFive we may face several
disconnected equivalence classes. In fact, for this reason, \SFiveU is
considered as metaphysically more appropriate by some philosophers;
cf. \cite[p.~127]{williamson13}.

Furthermore, for \SFiveU an improved embedding is possible. Universality implies
that the guarding predicates in the definitions of $\Box$ and
$\Diamond$ always hold. Therefore, they can be omitted and the
accessibility relation can be dispensed altogether. The modal
operators can then be defined merely as:
% \begin{itemize}
% \item $\BlackBox \varphi \equiv \forall w. (\varphi w)$ 
% \item $\BlackDia \varphi \equiv \exists w. (\varphi w)$
% \end{itemize}
\[\Box \varphi \equiv \lambda w. \forall v.  \varphi(v) \qquad \text{and}
\qquad \Dia \varphi \equiv \lambda w.\exists v. \varphi(v)\]

The new embedding of \SFiveU in Isabelle/HOL is shown in
Fig.~\ref{QML_S5U}.
With this improved embedding, the final theorem T3
(\textit{Necessarily, there exists God}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2.5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Wider adoption of HOL theorem proving technology for
reasoning about and within embedded object logics, especially among
non-expert users, is still hindered by the gap between the syntax used
by people, when they write logical formulas with pen and paper, and
the syntax used by HOL theorem provers. Even when the syntax of the
underlying higher-order system is elegant (as is the case in
Isabelle/HOL), the embedding of HOML into HOL may easily expose
details of HOL that may be uncommon to the user, disturbing his/her
experience while using the system. To illustrate this point,
Fig.~\ref{UglyEssence} shows how the definition of essence looked like
in previous work \cite{J28}, where advanced syntax-sugaring features
were not used. It looks notably higher-order, and its style differs
significantly from the common style seen in works on modal logics and
the ontological argument. The following specific issues can be
enumerated:
\begin{enumerate}
\item $\lambda$-abstractions, which are typically a HOL feature, appear explicitly in places where they did not need to in a pure HOML formulation (cf. G\"odel's manuscript, Fig.~\ref{GoedelScript}).
\item Quantifiers appear as higher-order defined constants, and not as binders. This forces the user to read (and write) formulas of the form $\forall (\lambda x. A(x))$ instead of the more common $\forall x. A(x)$.
\item The lifted modal connectives are represented by prefixing the
  letter ``m'' (e.g. $m\wedge$ and $m\imp$). The prefix disturbs the
  user, as it constantly reminds him/her that there is something
  unusual about the modal connectives.
\item Higher-order parenthesis conventions for the application of a predicate to a term are used. 
Instead of reading $\psi(y)$, as he/she would expect, he/she has to read $(\psi ~ y)$. Outside niche areas in computer science, the former syntax is more widely known than the latter.
\end{enumerate}

@@

On the other hand, there is still room for many pragmatical
improvements in Isabelle; just one example: in default setting,
Sledgehammer does not immediately inform the user when a proof has
been found and instead silently first executes a series of
time-consuming proof analysis processes (e.g. its dependency
minimization), before it eventually reports success. For G\"odel's
theorem T3 (\textit{Necessarily, there exists God}), for example, this
phase of silence takes several minutes --- during which the user might
actually give up on the proof attempt --- even though \textsc{Leo-II} already
reported success to Sledgehammer after 2.5 seconds.

@@

%\paragraph*{Acknowledgments} Will be added at a later point. ToDo
